---
title: "Introduction to padlocdev"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to padlocdev}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = TRUE
)
```

```{r setup, include = FALSE}
library(padlocdev)
```

The goal of padlocdev is to simplify the validation of padloc-db data, 
ensuring that the system models, HMMs, and metadata will work together as 
expected when running padloc.

### Vignettes

Explore the available vignettes for more information on the capabilities of 
padlocdev and typical workflows for processing padloc-db data.

| Vignette                      | Description |
|:------------------------------|:----------------------------------------------|
| `vignette("reading-data")`    | Learn how to read padloc-db data into R       |
| `vignette("validating-data")` | Learn how to validate padloc-db data          |
| `vignette("preparing-data")`  | Learn how to prepare padloc-db data for use   |
| `vignette("workflow")`        | A summarised workflow of padloc-db validation | 

### Example files

This package includes example files for demonstrating functions. These files
can be listed and accessed with the function `padlocdev_example()`. In the
following code, you'll see this function used to access these example files.

If `file = ...` is not specified, `padlocdev_example()` lists all example files 
(these are located in the package directory `padlocdev/inst/exdata/`).

```{r, message = FALSE}
padlocdev_example()
```

Otherwise, the full path to the file is generated.

```{r, message = FALSE}
padlocdev_example(file = "padloc-db/hmm/PDLC00150.hmm")
```
