---
title: "Preparing padloc-db data for use"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preparing padloc-db data for use}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = TRUE
)
```

```{r setup}
library(padlocdev)
```

### Spliting a database into groups

Once a padloc-db is validated, it is useful to split it into groups for
batched searches with padloc.

Read `sys_groups.txt` using `read_sys_groups()`.

```{r, message = FALSE}
path <- padlocdev_example("sys_groups.txt")
sys_groups <- read_sys_groups(path)
head(sys_groups)
```

Use `filter_models()` to filter a list of padloc models for those belonging
to a particular group, as defined in the system groups table.

```{r, message = FALSE}
path <- padlocdev_example("sys")
models <- multi_read_padloc_model(path)
models_group_1 <- filter_models(models, sys_groups, "group_1")
names(models_group_1)
```

To pull 

```{r, message = FALSE}
hmm_meta <- padlocdev_example("hmm_meta.txt") |> read_hmm_meta()
hmm_meta_expanded <- expand_hmm_meta(hmm_meta)
hmm_meta_group_1 <- filter_hmm_meta(models_group_1, hmm_meta_expanded)
```

To copy HMMs from a particular system group to a new directory use 
`system_cp_hmm()`. This, and related `system_cp_*` functions, use the `?fs` 
package to manipulate system files.

Before trying to copy any HMMs, it's important to verify that the name of the 
HMM file corresponds with the accession i.e. `ACC` field in the HMM header. 
This is not important for function of the database, though it is probably good 
practice, and it is important for this exercise as the accession is what we'll 
be using to identify HMMs to copy.

Use `verify_hmm_names()` to check that the names match. It returns a `list` of 
two `tibbles`, `names_match` which lists HMMs with matching file and accession 
names, and `names_mismatch` which lists HMMs with different file and accession 
names.

```{r, message = FALSE}
group_1_names_verify <- verify_hmm_names(models_group_1)
group_1_names_verify
```

In this case all names match, so we can continue. Use `filter_hmm_meta()` to 
extract rows from the HMM metadata table that are actually relevant to 

```{r, message = FALSE}
system_cp_hmm()
```

Use the wrapper `divide_database()` to create sub-databases for all groups in
the system groups table.

```{r, message = FALSE}
 # divide_database(
 #   sys_expanded = sys_expanded,
 #   sys_groups = sys_groups,
 #   hmm_meta_expanded = hmm_meta_expanded,
 #   sys_meta = sys_meta
 #   path = "path/to/database",
 #   new_path = "/path/to/new/database"
 # )
```
# 
# ```{r, message = FALSE}
# 
# ```
# 
# ```{r, message = FALSE}
# 
# ```
# 
# ```{r, message = FALSE}
# 
# ```
