---
title: "Reading padloc-db data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{reading-data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = TRUE
)
```

```{r setup}
library(padlocdev)
```

### Reading padloc-db files

You can read single profile HMMs, system models, the HMM metadata table, 
system metadata table, and system group information with the following 
functions.

#### Metadata tables

Read `hmm_meta.txt` with `read_hmm_meta()`.

```{r, message = FALSE}
path <- padlocdev_example("hmm_meta.txt")
hmm_meta <- read_hmm_meta(path)
head(hmm_meta)
```

Read `sys_meta.txt` with `read_sys_meta()`.

```{r, message = FALSE}
path <- padlocdev_example("sys_meta.txt")
sys_meta <- read_sys_meta(path)
head(sys_meta)
```

Read `sys_groups.txt` with `read_sys_groups()`.

```{r, message = FALSE}
path <- padlocdev_example("sys_groups.txt")
sys_groups <- read_sys_groups(path)
head(sys_groups)
```

All metadata tables are stored as `tibble` objects, and can be operated-on the
same as any other `tibble` (see `?tibble::tbl_df` for more information).

#### Profile HMMs

Read a full profile HMM, including header and model information with 
`read_hmm()`.

```{r, message = FALSE}
path <- padlocdev_example("hmm/PROT00001.hmm")
hmm <- read_hmm(path)
```

Profile HMMs are stored as `list` objects with two elements, `header` and 
`model`. 

```{r, message = FALSE}
names(hmm)
```

```{r, message = FALSE}
names(hmm[["header"]])
```

Accessing elements works the same as with any other list.

```{r, message = FALSE}
hmm[["header"]][["NAME"]]
# Or
hmm$header$NAME
# Or
purrr::pluck(hmm, "header", "NAME")
# Etc.
```

To avoid reading the model information, which can be quite large depending on
the size of the profile HMM, use `read_hmm_header()` instead. Currently, there
is no use for reading in the model information anyway, at least for the purpose
of data validation.

```{r, message = FALSE}
hmm_header <- read_hmm_header(path)
```

The object returned by `read_hmm_header()` is equivalent to subsetting the 
header element from the object returned by `read_hmm()` (i.e.
`read_hmm(path)[["header"]]`).

```{r, message = FALSE}
names(hmm_header)
```

#### System models

Read system models with `read_padloc_model()`.

```{r, message = FALSE}
path <- padlocdev_example("sys/SystemA_I.yaml")
model <- read_padloc_model(path)
```

System models are also stored as `list` objects with each element corresponding 
to a field in the original model file.

```{r, message = FALSE}
names(model)
```

```{r, message = FALSE}
model[["core_genes"]]
```

#### Reading multiple files

Wrappers are provided for reading all of the system model or profile HMM files 
in a directory into a list.

Use `multi_read_padloc_model()` to read all system models.

```{r, message = FALSE}
path <- padlocdev_example("sys")
models <- multi_read_padloc_model(path)
names(models)
```

```{r, message = FALSE}
models[["SystemA_I"]]
```

Use `multi_read_hmm()` to read all profile HMMs.

```{r, message = FALSE}
path <- padlocdev_example("hmm")
hmms <- multi_read_hmm(path)
names(hmms)
```

For larger databases, it's more efficient to just read the header information
of each HMM.

```{r, message = FALSE}
hmm_headers <- multi_read_hmm_header(path)
```

Be aware that, as before, the list output from `multi_read_hmm_header()` has 
one less level than that of `multi_read_hmm()`, since there's no model 
information.

```{r, message = FALSE}
names(hmms[["PROT00001"]][["header"]])
```

```{r, message = FALSE}
names(hmm_headers[["PROT00001"]])
```
