---
title: "Summarised workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Summarised workflow}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r debug, include = FALSE}
path_hmm <- "~/tools/padloc_update_230119/hmm"
path_sys <- "~/tools/padloc_update_230119/sys"
path_hmm_meta <- "~/tools/padloc_update_230119/hmm_meta.txt"
path_sys_meta <- "~/tools/padloc_update_230119/sys_meta.txt"
path_sys_groups <- "~/tools/padloc_update_230119/sys_groups.txt"
```

Load `padlocdev`.

```{r setup}
library(padlocdev)
```

Establish the paths to each database component.

```{r set_paths}
path_hmm <- "/path/to/padloc-db/hmm/"
path_sys <- "/path/to/padloc-db/sys/"
path_hmm_meta <- "/path/to/padloc-db/hmm_meta.txt"
path_sys_meta <- "/path/to/padloc-db/sys_meta.txt"
path_sys_groups <- "/path/to/padloc-db/sys_groups.txt"
```

Read-in the database data.

```{r read_data}
hmms <- multi_read_hmm_header(path_hmm)
models <- multi_read_padloc_model(path_sys)
hmm_meta <- read_hmm_meta(path_hmm_meta)
sys_meta <- read_sys_meta(path_sys_meta)
sys_groups <- read_sys_groups(path_sys_groups)
```

Verify that HMM file names match their accession.

```{r message = FALSE}
name_verification <- verify_hmm_names(hmms)
name_verification
```

Compare the profile HMMs with the HMM metadata table.

```{r message = FALSE}
hmm_files_hmm_meta_comparison <- compare_hmm_files_hmm_meta(hmms, hmm_meta)
hmm_files_hmm_meta_comparison
```

Compare the system models with the system metadata table.

```{r message = FALSE}
sys_files_sys_meta_comparison <- compare_sys_files_sys_meta(models, sys_meta)
sys_files_sys_meta_comparison
```

Compare the system metadata table with the system groups table.

```{r message = FALSE}
sys_meta_sys_comparison <- compare_sys_meta_sys_groups(sys_meta, sys_groups)
sys_meta_sys_comparison
```

Expand the HMM metadata table to account for HMMs with multiple name 
assignments, and expand the system models to account for groups of genes.

```{r preprocess_data}
hmm_meta_expanded <- expand_hmm_meta(hmm_meta)
models_expanded <- expand_gene_groups_all(models, hmm_meta_expanded)
```

Compare the system models with the HMM metadata table.

```{r}
sys_files_hmm_meta_comparison <- compare_sys_files_hmm_meta(models_expanded, hmm_meta_expanded)
sys_files_hmm_meta_comparison
```

Generate a validity report for the system models. Inspect the validity report 
to confirm that all models are valid. Update the models where necessary, then 
re-read and re-validate to confirm.

```{r validate_models}
validity_report <- report_padloc_model_validity(models_expanded)
why_invalid(validity_report)
```

Once everything is in order, divide the database into sub-groups for testing.

```{r postprocess_data}
divide_database(
  models_expanded, 
  sys_groups, 
  hmm_meta_expanded, 
  sys_meta, 
  "/path/to/origin/database", 
  "/path/to/sub/databases"
)
```

